---
title: "BST260 Code and Plots"
output: github_document
date: '2022-12-15'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(tidyverse)
library(data.table)
library(haven)
library(dslabs)
library(dplyr)
library(plyr)
library(ggplot2)

```

## Read in Behavioral Risk Factor Surveillance System data from 2010 

```{r}
brfss10 <- read_xpt("/Users/kierantodd/Desktop/CDBRFS10.XPT",
                    col_select = c(`IYEAR`,`_STATE`, `STOPSMK2`)) %>%
  as.data.table() %>%
  setnames(old = c("IYEAR","_STATE", "STOPSMK2"), new = c("YEAR","STATE", "MENTHLTH"))
```

## Read in Behavioral Risk Factor Surveillance System data from 2020 

```{r}
brfss20 <- read_xpt("/Users/kierantodd/Desktop/LLCP2020.XPT",
                    col_select = c(`IYEAR`,`_STATE`, `STOPSMK2`)) %>%
  as.data.table() %>%
  setnames(old = c("IYEAR","_STATE", "STOPSMK2"), new = c("YEAR","STATE", "MENTHLTH"))
```

##Combine BRFSS 2010 and BRFSS 2020 mental health data by State 

```{r}

brfss_all <- rbindlist(list(brfss10, brfss20), use.names = T)
brfss_all <- na.omit(brfss_all)

```

##Filter by Nevada and Arizona for the simple 2x2 difference-in-differences model 

```{r}
brfss_all <- brfss_all |> 
  filter(STATE %in% c("5", "32")) |>
  select(STATE, YEAR, MENTHLTH) %>%
  mutate(outcome = case_when(
    MENTHLTH == 88 ~ 0,
    MENTHLTH == 77 ~ NA_real_,
    MENTHLTH == 99 ~ NA_real_,
    TRUE ~ MENTHLTH))

na.omit(brfss_all)
```


##Rename States by their FIPS codes to the State Abbreviations
```{r}
brfss_all$STATE = ifelse(brfss_all$"STATE" == 5,"Arizona","Nevada")
```


##Further clean data so that only 2010 and 2020 interview dates exist, also omit any rows that have no values
```{r}
brfss_analysis <- brfss_all %>% filter(YEAR %in% c(2010,2020))
brfss_analysis <- na.omit(brfss_analysis)
```

##run a simple 2x2 difference in difference model using linear probablility model 
```{r}
did_model1 <- lm(outcome ~ factor(STATE)*factor(YEAR), data=brfss_analysis)
summary(did_model1)
```

# Plotting a density plot of mental health bad days in the pre period
```{r}
ggplot(data=brfss_analysis[brfss_analysis$YEAR==2010], aes(x=outcome, group=as.factor(STATE)))+
  scale_x_continuous(breaks = seq(from = 0, to = 30, by = 5)) +
  geom_density( aes( fill=as.factor(STATE) ), alpha=0.4 ) +
  theme_bw() + xlab("Number of Days") +
  ylab("") + guides(fill=guide_legend(title="State")) +
  ggtitle("Number of Bad Mental Health Days (in a month) in 2010")
```

# Plotting a density plot of mental health bad days in the post period (2020)
```{r}
ggplot(data=brfss_analysis[brfss_analysis$YEAR==2020], aes(x=outcome, group=as.factor(STATE)))+
  scale_x_continuous(breaks = seq(from = 0, to = 30, by = 5)) +
  geom_density( aes( fill=as.factor(STATE) ), alpha=0.4 ) +
  theme_bw() + xlab("Number of Days") +
  ylab("") + guides(fill=guide_legend(title="State")) +
  ggtitle("Number of Bad Mental Health Days (in a month) in 2020")
```

# Creating a plot to test for parallel pre-trends

```{r}

agg <- ddply(brfss_analysis,.(YEAR, STATE),summarize, val = mean(outcome))

ggplot( data=agg, aes(x=as.factor(YEAR), y=val, group=STATE,
                      color=as.factor(STATE))) +
  geom_line(lwd=1.5)+
  theme_bw() + geom_point(size=4) +
  ylab("Bad Mental Health Days") + xlab("Year") +
  guides(color=guide_legend(title="States")) +
  ggtitle("Mental Health Trajectories for Arizona v. Nevada")

```
