---
title: "Results"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(tidyverse)
library(data.table)
library(haven)
library(dslabs)
library(dplyr)
library(plyr)
library(ggplot2)




brfss10 <- read_xpt("/Users/kierantodd/Desktop/CDBRFS10.XPT",
                    col_select = c(`IYEAR`,`_STATE`, `MENTHLTH`)) %>%
  as.data.table() %>%
  setnames(old = c("IYEAR","_STATE", "MENTHLTH"), new = c("YEAR","STATE", "MENTHLTH"))





brfss20 <- read_xpt("/Users/kierantodd/Desktop/LLCP2020.XPT",
                    col_select = c(`IYEAR`,`_STATE`, `MENTHLTH`)) %>%
  as.data.table() %>%
  setnames(old = c("IYEAR","_STATE", "MENTHLTH"), new = c("YEAR","STATE", "MENTHLTH"))



brfss_all <- rbindlist(list(brfss10, brfss20), use.names = T)
brfss_all <- na.omit(brfss_all)

brfss_all <- brfss_all |> 
  filter(STATE %in% c("5", "32")) |>
  select(STATE, YEAR, MENTHLTH) %>%
  mutate(outcome = case_when(
    MENTHLTH == 88 ~ 0,
    MENTHLTH == 77 ~ NA_real_,
    MENTHLTH == 99 ~ NA_real_,
    TRUE ~ MENTHLTH))

na.omit(brfss_all)



brfss_all$STATE = ifelse(brfss_all$"STATE" == 5,"Arizona","Nevada")

brfss_analysis <- brfss_all %>% filter(YEAR %in% c(2010,2020))
brfss_analysis <- na.omit(brfss_analysis)
```


**Results**

Wrangling the data was necessary for the needs of the analytic technique chosen -- a difference-in-differences model. The data was originally filled with "n/a" observations, so removing any rows with "n/a" allowed for a complete data frame from which to work. Futher, combining the 2010 and 2020 BRFSS datasets into one dataframe was made easier by selecting only the columns that would be needed to conduct the simple difference-in-differences analysis. 

*The brfss 2010 data before wrangling*
```{r}
og_brfss10 <-read_xpt("/Users/kierantodd/Desktop/CDBRFS10.XPT")
head(og_brfss10)
```

*The BRFSS 2020 data before wrangling*
```{r}
og_brfss20 <-read_xpt("/Users/kierantodd/Desktop/LLCP2020.XPT")
head(og_brfss20)
                  

```
*Run a simple 2x2 difference in difference model using a linear probability model*
```{r}
did_model1 <- lm(outcome ~ factor(STATE)*factor(YEAR), data=brfss_analysis)
summary(did_model1)
```

Data from 15304 adults who participated in the BRFSS in 2010 and 2020 – 9167 adults from Arizona and 6137 adults from Nevada. In this simple 2x2 difference-in-difference analysis, being in was associated with a statistically significant increase in the number of bad mental health days among those who live in Nevada.


  In 2010, on average, the number of bad mental health days in Nevada was 1.04 days lower than in Arizona.  
  
  In 2020, on average, the number of bad mental health days in Nevada was 0.29 days less than the average number of bad mental health days in Nevada in 2010.
  
  The difference-in-difference estimate we can interpret as follows: in 2020, the number of bad mental health days among those in Nevada was, on average, an additional 0.63 days more than Arizona in the same year. If the difference-in-difference identifying assumptions hold, then this represents the causal effect of an increase in LGBTQ policy tally in Nevada relative to Arizona that maintained a low LGBTQ policy tally